---
title: "test"
output: html_document
date: "2024-11-11"
editor_options: 
  chunk_output_type: console
---
##Loading Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse);library(lubridate);library(here)
library(sf)
library(leaflet)
#install.packages("mapview")
library(mapview)
library(leafsync) #to display maps next to eachother
library(rvest)
library(dataRetrieval)
library(corrplot)

here()
```

##PM2.5 Data Wrangling

```{r PM2.5 visualization}

#spatial visualization of 2014 PM2.5 data
pm2.5_2014 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2014.csv')) 

pm2.5_sf_2014 <- pm2.5_2014 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2014, zcol = 'meanPM')

#spatial visualization of 2015 PM2.5 data
pm2.5_2015 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2015.csv')) 

pm2.5_sf_2015 <- pm2.5_2015 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2015, zcol = 'meanPM')

#spatial visualization of 2016 PM2.5 data
pm2.5_2016 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2016.csv')) 

pm2.5_sf_2016 <- pm2.5_2016 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2016, zcol = 'meanPM')

#spatial visualization of 2022 PM2.5 data
pm2.5_2017 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2017.csv')) 

pm2.5_sf_2017 <- pm2.5_2017 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2017, zcol = 'meanPM')

#spatial visualization of 2018 PM2.5 data
pm2.5_2018 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2018.csv')) 

pm2.5_sf_2018 <- pm2.5_2018 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#apview(pm2.5_sf_2018, zcol = 'meanPM')

#spatial visualization of 2019 PM2.5 data
pm2.5_2019 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2019.csv')) 

pm2.5_sf_2019 <- pm2.5_2019 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2019, zcol = 'meanPM')

#spatial visualization of 2020 PM2.5 data
pm2.5_2020 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2020.csv')) 

pm2.5_sf_2020 <- pm2.5_2020 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2020, zcol = 'meanPM')

#spatial visualization of 2021 PM2.5 data
pm2.5_2021 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2021.csv')) 

pm2.5_sf_2021 <- pm2.5_2021 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2021, zcol = 'meanPM')

#spatial visualization of 2022 PM2.5 data
pm2.5_2022 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2022.csv')) 

pm2.5_sf_2022 <- pm2.5_2022 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2022, zcol = 'meanPM')

#finding the two-year average in order to do analysis with EJ data

pm2.5_2013 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2013.csv')) #we need 2013
#data for the 2-year average at 2014

pm2.5_sf_2013 <- pm2.5_2013 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#combine spatial data frames for the 2-year averages
pm2.5_sf_20132014 <- rbind(pm2.5_sf_2013, pm2.5_sf_2014)

pm2.5_sf_20152016 <- rbind(pm2.5_sf_2015, pm2.5_sf_2016)

pm2.5_sf_20172018 <- rbind(pm2.5_sf_2017, pm2.5_sf_2018)

pm2.5_sf_20192020 <- rbind(pm2.5_sf_2019, pm2.5_sf_2020)

pm2.5_sf_20212022 <- rbind(pm2.5_sf_2021, pm2.5_sf_2022)

```

##Powerplant Data Wrangling
https://www.epa.gov/egrid/code-lookup

POWERPLANT Primary Fuel categories:
BIT - blast furnace gas -> most PM 2.5 producing

other PM2.5 producing types: 
AB - agricultural byproduct
BLQ - black liqour 
DFO - distillate fuel oil, light fuel oil, FO2, diesel oil
LFG - landfill gas
NG - natural gas
OBG - other biomass gas (digester gas, methane, and other biomass gases)
SLW - sludge waste 
WDS - wood, wood waste solid
WH - waste heat

excluded categories:
Not Available
SUN - solar
WND - wind
WAT - water
MWH - electricity 
NUC - nuclear

```{r read in powerplant data}
#do not think we should filter for this, should only be filtering for primary 
  #filter(TYPE != 'SOLAR PHOTOVOLTAIC') %>% 
  #filter(TYPE!='CONVENTIONAL HYDROELECTRIC') %>%
  #filter(TYPE!= 'BATTERIES; SOLAR PHOTOVOLTAIC') %>%
  #filter(TYPE!= 'NOT AVAILABLE')

powerplants <- st_read('https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Plants_gdb/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson') %>%
  filter(STATE == 'NC')


### BIT POWER PLANTS
  #major producer of PM 2.5 -> 18 BIT power plants in NC! 
  #combo of three types: 
     #conventional steam coal, conventional steam coal/natural gas fired 
     #combustion turbine, and conventional steam coal/petroleum liquids

BITfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL == 'BIT')


#other relatively high PM 2.5 producing from non-coal options
  #WDS, SLW, BLQ, AB, DFO
HighPMfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('WDS', 'SLW', 'BLQ', 'AB', 'DFO'))


#VERY low PM 2.5 emitting primary fuels!!! may be too small to even count
#LFG, NG, OBG, WAT, WH
otherPMproducingfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('LFG', 'NG', 'OBG', 'WH'))


#BIT plants and 5 other emitting ones
BIT_and_High_powerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('BIT', 'WDS', 'SLW', 'BLQ', 'AB', 'DFO' ))

```

```{r nc counties}
#nc counties shapefile
counties_sf<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>% 
  filter(STATEFP == 37)

```

```{r map power plant data mapview}

#mapview of BIT powerplants with NC counties
mapview(
  BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(counties_sf)

#mapview of BIT and pmpowerplants with NC counties
mapview(BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(HighPMfuelpowerplants,
          zcol = 'PRIM_FUEL') +
  mapview(counties_sf, col.region = 'grey') 

```

```{r operating or retired}
#maybe more of a correlation between retired vs operating?
operating_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'OP')
retired_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'RE')

#retirement_year contains all of the generators that retired in NC as of 
#Oct 2024 from https://www.eia.gov/electricity/data/eia860m/
retirement_year <- read.csv(here('Data/retired_generators.csv')) %>% 
  filter(Plant.State == "NC") %>% 
  group_by(Plant.Name, Plant.ID, Retirement.Year) %>% 
  # Keep Retirement.Year as the first, latest unit retirement is considered
  #retirement year
  summarise(Retirement.Year = first(Retirement.Year),.groups = "drop")

# Convert PLANT_CODE to integer in retired_plants
retired_plants <- retired_plants %>%
  mutate(PLANT_CODE = as.integer(PLANT_CODE)) 

# Perform the join, one plant ENKA is missing retirement year!
retired_plants <- retired_plants %>%
  left_join(retirement_year, by = c("PLANT_CODE" = "Plant.ID")) %>% 
  filter(Retirement.Year >= 2013)

mapview(operating_plants, zcol = "PRIM_FUEL", col.region = 'blue') +
  mapview(retired_plants, zcol = "PRIM_FUEL", col.region = 'red') +
  mapview(counties_sf, col.region = 'grey')

```

```{r visualizing plants by county}

#should merge pm data and then take means by county

countycount_OP_plants <- operating_plants %>%
  group_by(COUNTY, COUNTYFIPS) %>%
  summarize(Number_of_Plants = n())

countycount_RE_plants <- retired_plants %>%
  group_by(COUNTY, COUNTYFIPS, Retirement.Year) %>%
  summarize(Number_of_Plants = n())
```

##CDC Data Wrangling

```{r wrangling CDC data}
library(sf)

#wrangling 2014 data because 2014 gdb import does not work
SVI2014_csv <- read.csv(here('Data/Project_Data/NorthCarolina_county_2014.csv'), 
                        colClasses = c('FIPS' = 'factor'))
NC_counties <- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>% 
  filter(STATEFP == 37) %>% 
  rename(COUNTY = NAME)

SVI2014_gdb <- NC_counties %>% 
  left_join(
    SVI2014_csv, 
    by = c("GEOID" = "FIPS") 
    )

SVI2016_gdb <- st_read(here('Data/Project_Data/SVI2016_NORTHCAROLINA_COUNTY.gdb'))
SVI2018_gdb <- st_read(here('Data/Project_Data/SVI2018_NORTHCAROLINA_COUNTY.gdb'))
SVI2020_gdb <- st_read(here('Data/Project_Data/SVI2020_NORTHCAROLINA_COUNTY.gdb'))
SVI2022_gdb <- st_read(here('Data/Project_Data/SVI2022_NORTHCAROLINA_COUNTY.gdb'))

# Remove spatial information from SVI datasets
SVI2014_df <- SVI2014_gdb %>% st_drop_geometry() 
SVI2016_df <- SVI2016_gdb %>% st_drop_geometry()
SVI2018_df <- SVI2018_gdb %>% st_drop_geometry()
SVI2020_df <- SVI2020_gdb %>% st_drop_geometry()
SVI2022_df <- SVI2022_gdb %>% st_drop_geometry() 

#Renaming the GEOID column to FIPS to be able to join with other dataframes more 
#easily
SVI2014_df <- SVI2014_df %>%
  rename(FIPS = GEOID)

# Perform the joins on the selected columns for each of the themes to be able to 
#do GLM analysis for each theme
theme1 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME1), by = "FIPS")

colnames(theme1) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme2 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME2), by = "FIPS")

colnames(theme2) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme3 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME3), by = "FIPS")

colnames(theme3) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme4 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME4), by = "FIPS")

colnames(theme4) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme_agg <- SVI2014_df %>%
  select(FIPS, RPL_THEMES) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEMES), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEMES), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEMES), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEMES), by = "FIPS")

colnames(theme_agg) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)
```

##Spatial Visualization

```{r visualizing data}

counties_view <- mapview(counties_sf, col.region = 'grey')

plants_view <- mapview(operating_plants, zcol = "PRIM_FUEL", col.region = 'blue') +
  mapview(retired_plants, zcol = "PRIM_FUEL", col.region = 'red')

#theme 1 visualization
map_t1_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t1_20132014, map_t1_20152016, map_t1_20172018, map_t1_20192020, map_t1_20212022)

#theme 2 visualization
map_t2_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t2_20132014, map_t2_20152016, map_t2_20172018, map_t2_20192020, map_t2_20212022)

#theme 3 visualization
map_t3_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t3_20132014, map_t3_20152016, map_t3_20172018, map_t3_20192020, map_t3_20212022)

#theme 4 visualization
map_t4_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t4_20132014, map_t4_20152016, map_t4_20172018, map_t4_20192020, map_t4_20212022)

#Overall percentile visualization
map_overall_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_overall_20132014, map_overall_20152016, map_overall_20172018, map_overall_20192020, map_overall_20212022)
```

##GLMs/Linear Regression

```{r Data Wrangling for Linear Regression}

#add column titled 'FIPS' to all the PM2.5 data frames combining existing columns
#titled 'State FIPS Code' and 'County 'FIPS Code', with the goal of then
#combining with the SVI data frames which have 'FIPS' columns already

pm2.5_2013 <- pm2.5_2013 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`)) 

pm2.5_2014 <- pm2.5_2014 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2015 <- pm2.5_2015 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2016 <- pm2.5_2016 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2017 <- pm2.5_2017 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2018 <- pm2.5_2018 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2019 <- pm2.5_2019 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2020 <- pm2.5_2020 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2021 <- pm2.5_2021 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2022 <- pm2.5_2022 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

#combine PM2.5 data frames for 2-year averages once again

pm2.5_20132014 <- rbind(pm2.5_2013, pm2.5_2014) %>% 
  group_by(`State FIPS Code`, `County FIPS Code`, `County`, `FIPS`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    )
pm2.5_20152016 <- rbind(pm2.5_2015, pm2.5_2016) %>% 
  group_by(`State FIPS Code`, `County FIPS Code`, `County`, `FIPS`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    )
pm2.5_20172018 <- rbind(pm2.5_2017, pm2.5_2018) %>% 
  group_by(`State FIPS Code`, `County FIPS Code`, `County`, `FIPS`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    )
pm2.5_20192020 <- rbind(pm2.5_2019, pm2.5_2020) %>% 
  group_by(`State FIPS Code`, `County FIPS Code`, `County`, `FIPS`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    )
pm2.5_20212022 <- rbind(pm2.5_2021, pm2.5_2022) %>% 
  group_by(`State FIPS Code`, `County FIPS Code`, `County`, `FIPS`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    )

# Add year labels to each data frame
pm2.5_20132014$year <- 2014
pm2.5_20152016$year <- 2016
pm2.5_20172018$year <- 2018
pm2.5_20192020$year <- 2020
pm2.5_20212022$year <- 2022

# Combine the data frames into one
pm2.5_all <- rbind(pm2.5_20132014, pm2.5_20152016, pm2.5_20172018, 
                   pm2.5_20192020, pm2.5_20212022)

# Merge operating plant counts
pm2.5_all <- pm2.5_all %>%
  mutate(County_lower = tolower(`County`))  # Add a temporary lowercase column
pm2.5_all <- pm2.5_all %>%
   left_join(
    countycount_OP_plants %>% mutate(COUNTY_lower = tolower(COUNTY)),
    by = c("County_lower" = "COUNTY_lower"))

#Based on countycount_RE_plants df, manually do the following:
#for Chatham (add 4 for 2014)
pm2.5_all$Number_of_Plants[6] <- 4

#for New Hanover (add 8 for 2014, add 4 for 2016 and 2018)
pm2.5_all$Number_of_Plants[26] <- 8
pm2.5_all$Number_of_Plants[57] <- 4
pm2.5_all$Number_of_Plants[81] <- 4

#for Gaston (add 4 for 2014)
pm2.5_all$Number_of_Plants[13] <- pm2.5_all$Number_of_Plants[13] + 4

#for Bladen (add 1 for 2014, 2016, 2018) - no PM2.5 monitor so did not add

#for Edgecombe (add 1 for 2014, 2016, 2018, 2020)
pm2.5_all$Number_of_Plants[11] <- 1
pm2.5_all$Number_of_Plants[37] <- 1
pm2.5_all$Number_of_Plants[70] <- 1
pm2.5_all$Number_of_Plants[91] <- 1

# Reshape theme_agg from wide to long format
theme_agg_long <- theme_agg %>%
  pivot_longer(
    cols = -FIPS, # Keep FIPS as is; all other columns (years) will be reshaped
    names_to = "year", # Create a new column named 'year' from the wide column headers
    values_to = "theme_agg_value" # Create a column for the values
  )

theme1_long <- theme1 %>%
  pivot_longer(
    cols = -FIPS, # Keep FIPS as is; all other columns (years) will be reshaped
    names_to = "year", # Create a new column named 'year' from the wide column headers
    values_to = "theme1_value" # Create a column for the values
  )

theme2_long <- theme2 %>%
  pivot_longer(
    cols = -FIPS, # Keep FIPS as is; all other columns (years) will be reshaped
    names_to = "year", # Create a new column named 'year' from the wide column headers
    values_to = "theme2_value" # Create a column for the values
  )

theme3_long <- theme3 %>%
  pivot_longer(
    cols = -FIPS, # Keep FIPS as is; all other columns (years) will be reshaped
    names_to = "year", # Create a new column named 'year' from the wide column headers
    values_to = "theme3_value" # Create a column for the values
  )

theme4_long <- theme4 %>%
  pivot_longer(
    cols = -FIPS, # Keep FIPS as is; all other columns (years) will be reshaped
    names_to = "year", # Create a new column named 'year' from the wide column headers
    values_to = "them4_value" # Create a column for the values
  )

# Ensure the 'year' column is numeric to match pm2.5_all
theme_agg_long$year <- as.numeric(theme2_long$year)
theme1_long$year <- as.numeric(theme1_long$year)
theme2_long$year <- as.numeric(theme2_long$year)
theme3_long$year <- as.numeric(theme3_long$year)
theme4_long$year <- as.numeric(theme4_long$year)

# Ensure FIPS is consistent in both datasets (character type)
pm2.5_all$COUNTYFIPS <- as.character(pm2.5_all$COUNTYFIPS)
theme_agg_long$FIPS <- as.character(theme_agg_long$FIPS)
theme1_long$FIPS <- as.character(theme1_long$FIPS)
theme2_long$FIPS <- as.character(theme2_long$FIPS)
theme3_long$FIPS <- as.character(theme3_long$FIPS)
theme4_long$FIPS <- as.character(theme4_long$FIPS)

# Rename FIPS in theme dataframes to COUNTYFIPS
#colnames(theme_agg_long)[colnames(theme_agg_long) == "FIPS"] <- "COUNTYFIPS"
#colnames(theme1_long)[colnames(theme1_long) == "FIPS"] <- "COUNTYFIPS"
#colnames(theme2_long)[colnames(theme2_long) == "FIPS"] <- "COUNTYFIPS"
#colnames(theme3_long)[colnames(theme3_long) == "FIPS"] <- "COUNTYFIPS"
#colnames(theme4_long)[colnames(theme4_long) == "FIPS"] <- "COUNTYFIPS"

# Merge pm2.5_all with reshaped theme dataframes
combined_data <- merge(pm2.5_all, theme_agg_long, by = c("FIPS", "year"), 
                       all.x = TRUE)
combined_data <- merge(combined_data, theme1_long, by = c("FIPS", "year"), 
                       all.x = TRUE)
combined_data <- merge(combined_data, theme2_long, by = c("FIPS", "year"), 
                       all.x = TRUE)
combined_data <- merge(combined_data, theme3_long, by = c("FIPS", "year"), 
                       all.x = TRUE)
combined_data <- merge(combined_data, theme4_long, by = c("FIPS", "year"), 
                       all.x = TRUE)

#cleaning up combined_data
combined_data <- combined_data %>%
  select(-c(County_lower, COUNTY, COUNTYFIPS))
```

```{r Linear Regression}
#Correlation Plots: Exploratory plots of several continuous data points to 
#determine possible relationships, as well as covariance among explanatory 
#variables
combined_selectedcolumns <- 
  combined_data %>%
  select(meanPM, theme_agg_value, Number_of_Plants, theme1_value, theme2_value, 
         theme3_value, them4_value) %>%
  na.omit()
combined_data_corplot <- cor(combined_selectedcolumns)
corrplot(combined_data_corplot, method = "ellipse")

#Analysis of corrplot: When looking at the Correlation Ellipse Plot, we are looking at (i) the elliptical shape, (ii) the direction/slope of the ellipse, and (iii) the intensity of the ellipse's color to examine pairwise correlations. According to our correlation plot, the strongest correlations across 2014 to 2022 appear for the following variable pairs: (a) the aggregate of all four SVI themes and SVI theme 1, and (b) the aggregate of all four SVI themes and SVI theme 4, since the narrower ellipse shapes and darker ellipse colors suggest a very high correlation among these pairs. This makes sense since SVI theme 1 is one of the four themes that make up the SVI aggregate. Although the elliptical shapes and color intensities suggest strong correlations between the other SVI themes, individually, and the SVI aggregate, it is not as strong as the two pairs of variables first mentioned.

corrplot.mixed(combined_data_corplot, upper = "ellipse")

#Further correlation plot analysis: When looking at the mixed correlation plot, which combines the ellipses with the exact correlation values, we can confirm our prior observations when we see the correlation values for (a) the SVI aggregate and SVI theme 1 and (b) the SVI aggregate and SVI theme 4 are 0.85 and 0.78, respectively.


#run linear regression from 2014 to 2022 using PM2.5 as independent variable and 
#SVI measures and power plant count data as explanatory variables

multi.lin.reg <- lm(data = combined_data, meanPM ~ 
                        theme_agg_value + Number_of_Plants)

#AIC method to select variables to compute a stepwise regression that either 
#adds explanatory variables from the bottom up or removes explanatory variables 
#from a full set of suggested options. The smaller the AIC value, the better.
step(multi.lin.reg)
summary(multi.lin.reg)

#Analysis of multi linear regression/AIC: Using the Akaike Information Criterion (AIC) to evaluate the fit of a multiple regression model to our data, we see that the starting AIC value is 50.54. When removing the variable 'theme_agg_value', the AIC increases slightly. After then removing the variable 'Number_of_Plants', the AIC increases slightly once again. Because we want the smallest AIC value, this suggests that the inclusion of both these variables better explains the variation in mean PM2.5.

#However, a very low R^2 value, 0.08383, suggests that despite their inclusion, the SVI aggregate and number of power plants do not explain much of the variance in mean PM2.5 values from 2014 to 2022. Because the model only explains 8.38% of the data, the multi linear regression model may not provide a good fir for our data, and we should look to other factors that might better explain the variance in mean PM2.5 from 2014 to 2022.

#create 2 scatter plots: (i) PM2.5 data vs power plant count data and (ii) PM2.5 data vs SVI aggregate (all from 2014 to 2022)
#Scatterplot 1
pm2.5_powerplantcount <- 
  ggplot(combined_data, aes(x = Number_of_Plants, y = meanPM)) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  geom_point() 
print(pm2.5_powerplantcount)

#Scatterplot 2
pm2.5_SVIaggregate <- 
  ggplot(combined_data, aes(x = theme_agg_value, y = meanPM)) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  geom_point() 
print(pm2.5_SVIaggregate)

#Analysis of scatterplots: After plotting the relationships of (i) mean PM2.5 and the number of power plants from 2014 to 2022 and (ii) mean PM2.5 and the SVI aggregate from 2014 to 2022, the resulting scatterplots confirm that there appears to be no significant, or there is a very miniscule, relationship between PM2.5 and each of the number of power plants and the SVI aggregate for the years 2014 to 2022.

```


