---
title: "test"
output: html_document
date: "2024-11-11"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#hello
library(tidyverse);library(lubridate);library(here)
library(sf)
library(leaflet)
#install.packages("mapview")
library(mapview)
library(leafsync) #to display maps next to eachother
library(rvest)
library(dataRetrieval)

here()
```

```{r PM2.5 visualization}

#spatial visualization of 2014 PM2.5 data
pm2.5_2014 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2014.csv')) 

pm2.5_sf_2014 <- pm2.5_2014 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2014, zcol = 'meanPM')

#spatial visualization of 2015 PM2.5 data
pm2.5_2015 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2015.csv')) 

pm2.5_sf_2015 <- pm2.5_2015 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2015, zcol = 'meanPM')

#spatial visualization of 2016 PM2.5 data
pm2.5_2016 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2016.csv')) 

pm2.5_sf_2016 <- pm2.5_2016 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2016, zcol = 'meanPM')

#spatial visualization of 2022 PM2.5 data
pm2.5_2017 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2017.csv')) 

pm2.5_sf_2017 <- pm2.5_2017 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2017, zcol = 'meanPM')

#spatial visualization of 2018 PM2.5 data
pm2.5_2018 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2018.csv')) 

pm2.5_sf_2018 <- pm2.5_2018 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#apview(pm2.5_sf_2018, zcol = 'meanPM')

#spatial visualization of 2019 PM2.5 data
pm2.5_2019 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2019.csv')) 

pm2.5_sf_2019 <- pm2.5_2019 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2019, zcol = 'meanPM')

#spatial visualization of 2020 PM2.5 data
pm2.5_2020 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2020.csv')) 

pm2.5_sf_2020 <- pm2.5_2020 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2020, zcol = 'meanPM')

#spatial visualization of 2021 PM2.5 data
pm2.5_2021 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2021.csv')) 

pm2.5_sf_2021 <- pm2.5_2021 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2021, zcol = 'meanPM')

#spatial visualization of 2022 PM2.5 data
pm2.5_2022 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2022.csv')) 

pm2.5_sf_2022 <- pm2.5_2022 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#mapview(pm2.5_sf_2022, zcol = 'meanPM')

#finding the two-year average in order to do analysis with EJ data

pm2.5_2013 <- read_csv(here('Data/EPA_PM2.5/EPA_PM25_NC2013.csv')) #we need 2013
#data for the 2-year average at 2014

pm2.5_sf_2013 <- pm2.5_2013 %>% 
  group_by(`Site ID`, County, `Site Latitude`, `Site Longitude`) %>% 
  summarize(
    meanPM = mean(`Daily Mean PM2.5 Concentration`,na.rm=T),
    maxPM = max(`Daily Mean PM2.5 Concentration`,na.rm=T)
    ) %>% 
  st_as_sf(coords = c('Site Longitude','Site Latitude'), crs=4269)

#combine spatial data frames for the 2-year averages
pm2.5_sf_20132014 <- rbind(pm2.5_sf_2013, pm2.5_sf_2014)

pm2.5_sf_20152016 <- rbind(pm2.5_sf_2015, pm2.5_sf_2016)

pm2.5_sf_20172018 <- rbind(pm2.5_sf_2017, pm2.5_sf_2018)

pm2.5_sf_20192020 <- rbind(pm2.5_sf_2019, pm2.5_sf_2020)

pm2.5_sf_20212022 <- rbind(pm2.5_sf_2021, pm2.5_sf_2022)

```

https://www.epa.gov/egrid/code-lookup

POWERPLANT Primary Fuel categories:
BIT - blast furnace gas -> most PM 2.5 producing

other PM2.5 producing types: 
AB - agricultural byproduct
BLQ - black liqour 
DFO - distillate fuel oil, light fuel oil, FO2, diesel oil
LFG - landfill gas
NG - natural gas
OBG - other biomass gas (digester gas, methane, and other biomass gases)
SLW - sludge waste 
WDS - wood, wood waste solid
WH - waste heat

excluded categories:
Not Available
SUN - solar
WND - wind
WAT - water
MWH - electricity 
NUC - nuclear

```{r read in powerplant data}
#do not think we should filter for this, should only be filtering for primary 
  #filter(TYPE != 'SOLAR PHOTOVOLTAIC') %>% 
  #filter(TYPE!='CONVENTIONAL HYDROELECTRIC') %>%
  #filter(TYPE!= 'BATTERIES; SOLAR PHOTOVOLTAIC') %>%
  #filter(TYPE!= 'NOT AVAILABLE')

powerplants <- st_read('https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Plants_gdb/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson') %>%
  filter(STATE == 'NC')


### BIT POWER PLANTS
  #major producer of PM 2.5 -> 18 BIT power plants in NC! 
  #combo of three types: 
     #conventional steam coal, conventional steam coal/natural gas fired 
     #combustion turbine, and conventional steam coal/petroleum liquids

BITfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL == 'BIT')


#other relatively high PM 2.5 producing from non-coal options
  #WDS, SLW, BLQ, AB, DFO
HighPMfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('WDS', 'SLW', 'BLQ', 'AB', 'DFO'))


#VERY low PM 2.5 emitting primary fuels!!! may be too small to even count
#LFG, NG, OBG, WAT, WH
otherPMproducingfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('LFG', 'NG', 'OBG', 'WH'))


#BIT plants and 5 other emitting ones
BIT_and_High_powerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('BIT', 'WDS', 'SLW', 'BLQ', 'AB', 'DFO' ))

```

```{r nc counties}
#nc counties shapefile
counties_sf<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>% 
  filter(STATEFP == 37)

```

```{r map power plant data mapview}

#mapview of BIT powerplants with NC counties
mapview(
  BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(counties_sf)

#mapview of BIT and pmpowerplants with NC counties
mapview(BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(HighPMfuelpowerplants,
          zcol = 'PRIM_FUEL') +
  mapview(counties_sf, col.region = 'grey') 

```

```{r operating or retired}
#maybe more of a correlation between retired vs operating?
operating_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'OP')
retired_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'RE')

#retirement_year contains all of the generators that retired in NC as of 
#Oct 2024 from https://www.eia.gov/electricity/data/eia860m/
retirement_year <- read.csv(here('Data/retired_generators.csv')) %>% 
  filter(Plant.State == "NC") %>% 
  group_by(Plant.Name, Plant.ID, Retirement.Year) %>% 
  # Keep Retirement.Year as the first, latest unit retirement is considered
  #retirement year
  summarise(Retirement.Year = first(Retirement.Year),.groups = "drop")

# Convert PLANT_CODE to integer in retired_plants
retired_plants <- retired_plants %>%
  mutate(PLANT_CODE = as.integer(PLANT_CODE)) 

# Perform the join, one plant ENKA is missing retirement year!
retired_plants <- retired_plants %>%
  left_join(retirement_year, by = c("PLANT_CODE" = "Plant.ID")) %>% 
  filter(Retirement.Year >= 2013)

mapview(operating_plants, zcol = "PRIM_FUEL", col.region = 'blue') +
  mapview(retired_plants, zcol = "PRIM_FUEL", col.region = 'red') +
  mapview(counties_sf, col.region = 'grey')

```

```{r visualizing plants by county}

#should merge pm data and then take means by county
BITfuelpowerplants$COUNTY

BIT_and_High_powerplants$COUNTY

```

```{r wrangling CDC data}
library(sf)

#wrangling 2014 data because 2014 gdb import does not work
SVI2014_csv <- read.csv(here('Data/Project_Data/NorthCarolina_county_2014.csv'), 
                        colClasses = c('FIPS' = 'factor'))
NC_counties <- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>% 
  filter(STATEFP == 37) %>% 
  rename(COUNTY = NAME)

SVI2014_gdb <- NC_counties %>% 
  left_join(
    SVI2014_csv, 
    by = c("GEOID" = "FIPS") 
    )

SVI2016_gdb <- st_read(here('Data/Project_Data/SVI2016_NORTHCAROLINA_COUNTY.gdb'))
SVI2018_gdb <- st_read(here('Data/Project_Data/SVI2018_NORTHCAROLINA_COUNTY.gdb'))
SVI2020_gdb <- st_read(here('Data/Project_Data/SVI2020_NORTHCAROLINA_COUNTY.gdb'))
SVI2022_gdb <- st_read(here('Data/Project_Data/SVI2022_NORTHCAROLINA_COUNTY.gdb'))

# Remove spatial information from SVI datasets
SVI2014_df <- SVI2014_gdb %>% st_drop_geometry() 
SVI2016_df <- SVI2016_gdb %>% st_drop_geometry()
SVI2018_df <- SVI2018_gdb %>% st_drop_geometry()
SVI2020_df <- SVI2020_gdb %>% st_drop_geometry()
SVI2022_df <- SVI2022_gdb %>% st_drop_geometry() 

#Renaming the GEOID column to FIPS to be able to join with other dataframes more 
#easily
SVI2014_df <- SVI2014_df %>%
  rename(FIPS = GEOID)

# Perform the joins on the selected columns for each of the themes to be able to 
#do GLM analysis for each theme
theme1 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME1), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME1), by = "FIPS")

colnames(theme1) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme2 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME2), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME2), by = "FIPS")

colnames(theme2) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme3 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME3), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME3), by = "FIPS")

colnames(theme3) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

theme4 <- SVI2014_df %>%
  select(FIPS, RPL_THEME1) %>%
  inner_join(SVI2016_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2018_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2020_df %>% select(FIPS, RPL_THEME4), by = "FIPS") %>%
  inner_join(SVI2022_df %>% select(FIPS, RPL_THEME4), by = "FIPS")

colnames(theme4) <- c("FIPS", 2014, 2016, 2018, 2020, 2022)

#mapview(SVI2018_gdb, zcol = "RPL_THEME1") + mapview(powerplants, col.regions = "black") + mapview(pm2.5_sf, cex = "meanPM")
```

VISUALIZATION

```{r visualizing data}

counties_view <- mapview(counties_sf, col.region = 'grey')

plants_view <- mapview(operating_plants, zcol = "PRIM_FUEL", col.region = 'blue') +
  mapview(retired_plants, zcol = "PRIM_FUEL", col.region = 'red')

#theme 1 visualization
map_t1_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t1_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME1", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t1_20132014, map_t1_20152016, map_t1_20172018, map_t1_20192020, map_t1_20212022)

#theme 2 visualization
map_t2_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t2_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME2", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t2_20132014, map_t2_20152016, map_t2_20172018, map_t2_20192020, map_t2_20212022)

#theme 3 visualization
map_t3_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t3_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME3", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t3_20132014, map_t3_20152016, map_t3_20172018, map_t3_20192020, map_t3_20212022)

#theme 4 visualization
map_t4_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_t4_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEME4", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_t4_20132014, map_t4_20152016, map_t4_20172018, map_t4_20192020, map_t4_20212022)

#Overall percentile visualization
map_overall_20132014 <- mapview(pm2.5_sf_20132014, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2014_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20152016 <- mapview(pm2.5_sf_20152016, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2016_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20172018 <- mapview(pm2.5_sf_20172018, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2018_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20192020 <- mapview(pm2.5_sf_20192020, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2020_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

map_overall_20212022 <- mapview(pm2.5_sf_20212022, zcol = "meanPM", col.regions = colorRampPalette(c("lightgreen", "green", "darkgreen"))) + 
  mapview(SVI2022_gdb, zcol = "RPL_THEMES", 
          col.regions = colorRampPalette(c("pink", "orange", "purple"))) + 
  counties_view + plants_view

# Synchronize and display the maps side by side
sync(map_overall_20132014, map_overall_20152016, map_overall_20172018, map_overall_20192020, map_overall_20212022)
```

GLMs/Linear Regression

```{r Linear Regression}

#add column titled 'FIPS' to all the PM2.5 data frames combining existing columns
#titled 'State FIPS Code' and 'County 'FIPS Code', with the goal of then
#combining with the SVI data frames which have 'FIPS' columns already

pm2.5_2013 <- pm2.5_2013 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2014 <- pm2.5_2014 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2015 <- pm2.5_2015 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2016 <- pm2.5_2016 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2017 <- pm2.5_2017 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2018 <- pm2.5_2018 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2019 <- pm2.5_2019 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2020 <- pm2.5_2020 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2021 <- pm2.5_2021 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

pm2.5_2022 <- pm2.5_2022 %>%
  mutate(FIPS = paste0(`State FIPS Code`, `County FIPS Code`))

#combine PM2.5 data frames for 2-year averages once again

pm2.5_20132014 <- rbind(pm2.5_2013, pm2.5_2014)
pm2.5_20152016 <- rbind(pm2.5_2015, pm2.5_2016)
pm2.5_20172018 <- rbind(pm2.5_2017, pm2.5_2018)
pm2.5_20192020 <- rbind(pm2.5_2019, pm2.5_2020)
pm2.5_20212022 <- rbind(pm2.5_2021, pm2.5_2022)

#create data frames for 2-year intervals: 2014, 2016, 2018, 2020, and 2022 
#(using PM2.5 and SVI data)
combined_2014 <- merge(pm2.5_20132014, SVI2014_df, by = "FIPS", all = TRUE)

combined_2016 <- merge(pm2.5_20152016, SVI2016_df, by = "FIPS", all = TRUE)

combined_2018 <- merge(pm2.5_20172018, SVI2018_df, by = "FIPS", all = TRUE)

combined_2020 <- merge(pm2.5_20192020, SVI2020_df, by = "FIPS", all = TRUE)

combined_2022 <- merge(pm2.5_20212022, SVI2022_df, by = "FIPS", all = TRUE)

#run linear regression on years 2014, 2016, 2018, 2020, and 2022 using PM2.5
#as independent variable and SVI measures as explanatory variables

mlt.lin.reg2014 <- lm(data = combined_2014, `Daily Mean PM2.5 Concentration` ~ 
                        SPL_THEME1 + SPL_THEME2 + SPL_THEME3 + SPL_THEME4)

#Correlation Plots: Exploratory plots of several continuous data points to 
#determine possible relationships, as well as covariance among explanatory 
#variables
library(corrplot)
combined_2014 <- 
  combined_2014 %>%
  select(`Daily Mean PM2.5 Concentration`, SPL_THEME1, SPL_THEME2, SPL_THEME3, 
         SPL_THEME4) %>%
  na.omit()
combined_2014_corplot <- cor(combined_2014)
corrplot(combined_2014_corplot, method = "ellipse")
corrplot.mixed(combined_2014_corplot, upper = "ellipse")

#AIC method to select variables to compute a stepwise regression that either 
#adds explanatory variables from the bottom up or removes explanatory variables 
#from a full set of suggested options. The smaller the AIC value, the better.
step(mlt.lin.reg2014)

#create scatter plots with each of the four themes and compare for each year, 
#starting with the year 2014
pm2.5_2014_SVItheme1 <- 
  ggplot(combined_2014, aes(x = SPL_THEME4, y = `Daily Mean PM2.5 Concentration`)) +
  geom_smooth(method = "lm") +
  scale_y_log10() +
  geom_point() 
print(pm2.5_2014_SVItheme1)


```

