---
title: "Powerplantdata"
author: "Lauren Shohan"
date: "2024-11-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, warning=F, error=F}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse);library(here);
library(sf);library(mapview);library(leaflet)
here()
```

from lab 9 spatial analysis: 
describing the dataset we are using!!

Homeland Infrastructure Foundation-Level Data (HIFLD) Their open data site (<https://hifld-geoplatform.opendata.arcgis.com/>) has links to many datasets.

Navigate to <https://hifld-geoplatform.hub.arcgis.com/pages/hifld-open>
-   Scroll down to the Explore Categories area. Select Energy (for example)
-   Search for power plants.
-   Select the first return and open its [link](https://duke.is/HIFLD-PP)
-   Locate the APIs dropdown and copy the link to the GeoJSON option.

If you have difficulty, the link you want is: 
-   <https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Plants_gdb/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson>


My stuff: 
https://www.epa.gov/egrid/code-lookup

POWERPLANT Primary Fuel categories:
BIT - blast furnace gas -> most PM 2.5 producing

other PM2.5 producing types: 
AB - agricultural byproduct
BLQ - black liqour 
DFO - distillate fuel oil, light fuel oil, FO2, diesel oil
LFG - landfill gas
NG - natural gas
OBG - other biomass gas (digester gas, methane, and other biomass gases)
SLW - sludge waste 
WDS - wood, wood waste solid
WH - waste heat

excluded categories:
Not Available
SUN - solar
WND - wind
WAT - water
MWH - electricity 
NUC - nuclear


```{r read in powerplant data}
#do not think we should filter for this, should only be filtering for primary 
  #filter(TYPE != 'SOLAR PHOTOVOLTAIC') %>% 
  #filter(TYPE!='CONVENTIONAL HYDROELECTRIC') %>%
  #filter(TYPE!= 'BATTERIES; SOLAR PHOTOVOLTAIC') %>%
  #filter(TYPE!= 'NOT AVAILABLE')

powerplants <- st_read('https://services1.arcgis.com/Hp6G80Pky0om7QvQ/arcgis/rest/services/Plants_gdb/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson') %>%
  filter(STATE == 'NC')


### BIT POWER PLANTS
  #major producer of PM 2.5 -> 18 BIT power plants in NC! 
  #combo of three types: 
     #conventional steam coal, conventional steam coal/natural gas fired 
     #combustion turbine, and conventional steam coal/petroleum liquids

BITfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL == 'BIT')


#other relatively high PM 2.5 producing from non-coal options
  #WDS, SLW, BLQ, AB, DFO
HighPMfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('WDS', 'SLW', 'BLQ', 'AB', 'DFO'))


#VERY low PM 2.5 emitting primary fuels!!! may be too small to even count
#LFG, NG, OBG, WAT, WH
otherPMproducingfuelpowerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('LFG', 'NG', 'OBG', 'WH'))


#BIT plants and 5 other emitting ones
BIT_and_High_powerplants <- powerplants %>%
  filter(PRIM_FUEL %in% c('BIT', 'WDS', 'SLW', 'BLQ', 'AB', 'DFO' ))

```


```{r nc counties}
#nc counties shapefile
counties_sf<- st_read(here('Data/Spatial/cb_2018_us_county_20m.shp')) %>% 
  filter(STATEFP == 37)

```

```{r map power plant data mapview}

#mapview of BIT powerplants with NC counties
mapview(
  BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(counties_sf)

#mapview of BIT and pmpowerplants with NC counties
mapview(BITfuelpowerplants,
  zcol = 'PRIM_FUEL') +
  mapview(HighPMfuelpowerplants,
          zcol = 'PRIM_FUEL') +
  mapview(counties_sf, col.region = 'grey') 

```

```{r operating or retired}
#maybe more of a correlation between retired vs operating?
#not working well yet
operating_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'OP')
retired_plants <- BIT_and_High_powerplants %>% filter(STATUS == 'RE')

#retirement_year contains all of the generators that retired in NC as of 
#Oct 2024 from https://www.eia.gov/electricity/data/eia860m/
retirement_year <- read.csv(here('Data/retired_generators.csv')) %>% 
  filter(Plant.State == "NC") %>% 
  group_by(Plant.Name, Plant.ID, Retirement.Year) %>% 
  # Keep Retirement.Year as the first, latest unit retirement is considered
  #retirement year
  summarise(Retirement.Year = first(Retirement.Year),.groups = "drop")

# Convert PLANT_CODE to integer in retired_plants
retired_plants <- retired_plants %>%
  mutate(PLANT_CODE = as.integer(PLANT_CODE)) 

# Perform the join, one plant ENKA is missing retirement year!
retired_plants <- retired_plants %>%
  left_join(retirement_year, by = c("PLANT_CODE" = "Plant.ID")) #%>% 
  #filter(Retirement.Year >= 2013)

mapview(operating_plants, zcol = "PRIM_FUEL", col.region = 'blue') +
  mapview(retired_plants, zcol = "PRIM_FUEL", col.region = 'red') +
  mapview(counties_sf, col.region = 'grey')

```

```{r visualizing plants by county}

#should merge pm data and then take means by county
BITfuelpowerplants$COUNTY

BIT_and_High_powerplants$COUNTY

```


```
